Modularly Composing Typed Language Fragments

### 2014-11-04 att-pldi15/att-pldi15.tex
Updated

\AtBeginDocument{%
 \abovedisplayskip=5pt
 \abovedisplayshortskip=4pt
 \belowdisplayskip=4pt
 \belowdisplayshortskip=4pt
}

@@

In this paper, we begin from first principles by specifying a language, @$\lambda$, in the style of many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant on record types and constrained string types) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of its  term-level operators (e.g. record projection) using a static language where types and translations are values. The semantics includes a simple check that leads to powerful metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that  \emph{tycon invariants} will always be conserved under extension. Type system providers need not provide mechanized proofs. Instead, the check enforces a \emph{translation independence} property via type abstraction, the same principle underlying ML-style module systems.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation
\section{Introduction}\label{intro}
Typed programming languages are most often described as being composed from \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
In his textbook, Harper organizes such fragments around type constructors, describing each in a different chapter \cite{pfpl}. Languages are then identified by a set of these type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language of partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
In the literature, researchers often describe fragments using a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.

@@

Universality properties (e.g. ``Turing-completeness'') are often invoked to guarantee that an embedding that preserves a desirable fragment's dynamics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics. This is not always possible. 
Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a fragment must be exposed directly, rather than as a library, a new \emph{dialect} of a language must be constructed. Within the ML lineage, for example, dialects that go beyond ``core ML'' abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

@@

\item
\textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness or security guarantees, manage unwieldy lower-level abstractions and run-time systems or control cost are also frequently introduced in dialects, e.g. for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}. %\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.

\item
\textbf{Foreign Fragments:} A safe and natural foreign function interface (FFI) can be valuable (particularly given this proliferation of dialects). This requires enforcing the type system of the foreign language in the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
%Safe FFIs generally require direct extensions to the language. 
For example, MLj builds in a safe FFI to Java \cite{Benton:1999:IWW:317636.317791}.
\end{enumerate}
\vspace{-5px}

This \emph{dialect-oriented} state of affairs is unsatisfying. %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. 
While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both fragments may not be available. Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e. a safe FFI, item 3 above, is needed between every pair of dialects).

These problems do not arise for a fragment that can be expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism needs only to be established in the ``closed world'' of the module. For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g. uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client \cite{harper1997programming}. Other langauges offer similar facilities, often grouped under the moniker of \emph{abstract data types} \cite{liskov1974programming}. %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

@@

\paragraph{Contributions} In this paper, we take foundational steps towards this goal by  constructing a simple but surprisingly powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus). %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties. %a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR? Arxiv?}, reviewed in Sec. \ref{overview}. 
Its semantics are structured like those of many modern languages, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a much simpler \emph{internal language} (IL). 
Rather than building in a monolithic set of external type constructors, however, the  semantics are indexed by a \emph{tycon context}. Each tycon defines the semantics of its operators via functions written in a \emph{static language} (SL) where types and translations are values. %, which provides a form of type-level computation. %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).

@@

\section{Overview of @$\lambda$}\label{atlam}\label{overview}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-IL}
\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL). Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
\label{syntax-IL}
\vspace{-10px}
\end{figure}
\paragraph{External Language}
Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$. The abstract syntax of external terms is shown in Figure \ref{syntax-EL} (we will give various concrete desugarings in Sec. \ref{external-terms}). %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec. \ref{desugaring}. Our main focus here is on semantic extensions. 
The static and dynamic semantics are given simultaneously as a \emph{bidirectionally typed translation semantics}, i.e. the key judgements take the form:\[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent
These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax. In particular, note that the type is an ``output'' only in the synthetic judgement.

@@

External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$, where $\Gamma$ maps variables to the corresponding type translation.

@@

\paragraph{Internal Language} 
%Internal terms, $\iota$, together with internal types, $\tau$, form the \emph{typed internal language}. 
@$\lambda$ relies on a typed internal language supporting type abstraction (i.e. universal quantification over types). We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of any intermediate language for a typed functional language. %Note that the IL has a fixed semantics; changes to the IL induce different dialects of @$\lambda$.

We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}. 
%The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}. The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$. Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity.

@@

\paragraph{Static Language}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $n$ over natural numbers.}\vspace{-5px}
\label{syntax-SL}
\end{figure}
The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds (constrained by the standard positivity condition to prevent non-termination), and products and sums. The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq. The semantics also directly follow \cite{pfpl}, so we  omit the details of this core here.  Only three new  kinds are needed for the SL to serve its role as the language used to   control typechecking and translation of the EL: $\kty$ (Sec. \ref{types}), $\kity$ (Sec. \ref{sec:type-translations}) and $\kitm$ (Sec. \ref{sec:introop}).

@@

\inferrule[k-otherty]{
    \sofkX{\sttyidx}{\keyw{Nat} \times \kity}
}{
    \sofkX{\sty{\keyw{other}[m]}{\sttyidx}}{\kty}
}\vspace{-10px}
\end{mathpar}
\caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
\label{fig:types}\vspace{-8px}
\end{figure*}

@@

\paragraph{Examples}
Our first example of a user-defined tycon is $\tcvar{rstr}$. It has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way). Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which statically obey the invariant that they are in the regular language specified by the type index \cite{sanitation-psp14}. For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.  The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably in libraries \cite{TSLs}. We define the tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.\footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision. }

Our second example is the tycon $\tcvar{lprod}$, which will define a variant of labeled product type (labeled products are like record types, but maintain a row  ordering; record types are also definable in a manner discussed in the supplement, but maintaining an ordering simplifies our discussion). We choose the index kind of $\tcvar{lprod}$ to be $\klist{\klbl \times \kty}$, where list kinds are defined as inductive sums in the usual way, and $\klbl$ classifies static representations of row labels. The tycon context containing only $\tcvar{lprod}$'s definition is $\small\Phi_\text{lprod} := \tcdef{\tcvar{lprod}}{\tcsig{\klist{\klbl \times \kty}}{\chi_\text{lprod}}}{\theta_\text{lprod}}$. In the tycon context containing both tycon definitions, $\Phi_\text{rstr}\Phi_\text{lprod}$, we can define a labeled product type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$. Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and we again use kind-specific syntax, in this case for $\klbl$ and $\klist{\klbl \times \kty}$.

\subsection{Static Type Case Analysis}
\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
\small
\inferrule[tcc-ext]{
    \vdash \Phi\\
    \tc \notin \text{dom}(\Phi)\\
    \keq{\emptyset}{\ktyidx}\\
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
}{
    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
}\vspace{-10px}
\end{mathpar}
\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
\label{fig:tycon-ctxs}
\vspace{-8px}
\end{figure}

@@

The dynamics (see supplement) are straightforwardly consistent with the intuition above.% We will see an example of its use in Sec. \ref{sec:targops}.
\subsection{Tycon Context Well-Definedness}
The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.

\subsection{Type Equivalence}\label{sec:type-equivalence}
The first check simplifies the handling of type equivalence: type index kinds must be \emph{equality kinds}, i.e. those for which semantic equivalence implies syntactic equality at normal form. We define these by the judgement $\keq{\kDelta}{\kappa}$ (see supplement). 
Equality kinds are similar to equality types as found in Standard ML \cite{Tofte:89:TheDefinitionOfStandardML}. The main implication of this choice is that type indices cannot contain static functions.

@@

Applying the substitutions $\gamma_\text{ex/abs}$ and $\delta_\text{ex/abs}$ in the conclusion of the rule, we arrive at the actual term translation $\iota_\text{ex}$.

>>>>

\AtBeginDocument{%
 \abovedisplayskip=4pt
 \abovedisplayshortskip=4pt
 \belowdisplayskip=4pt
 \belowdisplayshortskip=4pt
}

\usepackage[compact]{titlesec}  
\titlespacing{\section}{0pt}{6pt}{2pt}
\titlespacing{\subsection}{0pt}{3pt}{0pt}
\titlespacing{\subsubsection}{0pt}{3pt}{0pt}
\titlespacing{\paragraph}{0pt}{3pt}{5pt}

@@

In this paper, we begin from first principles by specifying a language, @$\lambda$, in the style of many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show constrained strings and a variant on records) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of its  term-level operators (e.g. record projection) using a static language where types and translations are values. The semantics includes a simple check that leads to powerful metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that  \emph{tycon invariants} will always be conserved under extension. Type system providers need not provide mechanized proofs. Instead, the check enforces a \emph{translation independence} property via type abstraction, the same principle underlying ML-style module systems.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation
\section{Introduction}\label{intro}
Typed programming languages are most often described as being composed from \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
In his textbook, Harper organizes such fragments around type constructors, describing each in a different chapter \cite{pfpl}. Languages are then identified by a set of these type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language of partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
Another common practice is to describe a fragment using a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language (e.g. \cite{sanitation-psp14}).

@@

Universality properties (e.g. ``Turing-completeness'') are often invoked to guarantee that an embedding that preserves a desirable fragment's dynamics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics. This is not always possible. 
Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a fragment must be exposed directly, rather than as a library, a new \emph{dialect} of the language must be constructed. Within the ML lineage, for example, dialects that go beyond ``core ML'' abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

@@

\item
\textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness or security guarantees, manage unwieldy lower-level abstractions and run-time systems or control cost are also frequently introduced in dialects, e.g. for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}. %\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
\vspace{-3px}
\item
\textbf{Foreign Fragments:} A safe and natural foreign function interface (FFI) can be valuable (particularly given this proliferation of dialects). This requires enforcing the type system of the foreign language in the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
%Safe FFIs generally require direct extensions to the language. 
For example, MLj builds in a safe FFI to Java \cite{Benton:1999:IWW:317636.317791}.
\end{enumerate}
\vspace{-5px}

This \emph{dialect-oriented} state of affairs is unsatisfying. %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. 
While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both fragments may not be available. Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e. a safe FFI, item 3 above, is needed between every pair of dialects). And as just mentioned, composing dialects is non-trivial.

These problems do not arise for a fragment that can be expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism needs only to be established in the ``closed world'' of the module. For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g. uniqueness, if using a list representation) will hold no matter which other modules are in use by a client \cite{harper1997programming}. Other languages provide similar facilities, e.g. Scala's abstract type members \cite{conf/oopsla/AminRO14}. %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

@@

\paragraph{Contributions} In this paper, we take foundational steps toward this goal by  constructing a simple but surprisingly powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus). %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties. %a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR? Arxiv?}, reviewed in Sec. \ref{overview}. 
Its semantics are structured like those of many modern languages, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a much simpler \emph{internal language} (IL). 
Rather than building in a monolithic set of external type constructors, however, the  semantics are indexed by a \emph{tycon context}. Each tycon defines the semantics of its operators via functions written in a \emph{static language} (SL) where types and translations are values. %, which provides a form of type-level computation. %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).

@@

\section{Overview of @$\lambda$}\label{atlam}\label{overview}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-IL}
\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL). Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
\label{syntax-IL}
\vspace{-10px}
\end{figure}
\paragraph{External Language}
Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$. The abstract syntax of external terms is shown in Figure \ref{syntax-EL} (we will give various concrete desugarings in Sec. \ref{external-terms}). %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec. \ref{desugaring}. Our main focus here is on semantic extensions. 
The static and dynamic semantics are specified simultaneously as a \emph{bidirectionally typed translation semantics}, i.e. the key judgements take the form:\[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent
These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax. In particular, note that the type is an ``output'' only for the synthetic judgement.

@@

External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$ to construct a $\Gamma$ that maps the same variables to corresponding type translations.

@@

\paragraph{Internal Language} 
%Internal terms, $\iota$, together with internal types, $\tau$, form the \emph{typed internal language}. 
@$\lambda$ requires a typed internal language supporting type abstraction (i.e. universal quantification over types) \cite{Reynolds94anintroduction}. We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of a typical intermediate language for a typed functional language. %Note that the IL has a fixed semantics; changes to the IL induce different dialects of @$\lambda$.

We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}. 
%The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}. The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$. Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and omit the details.

@@

\paragraph{Static Language}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $n$ over natural numbers.}\vspace{-5px}
\label{syntax-SL}
\end{figure}
The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. The portion of the SL covered by the first row of kinds and first row of static terms, some of which are elided for concision, forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds, and products and sums. The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq and we will assume standard conveniences from such languages (e.g. let bindings, pattern matching and inference of type parameters when applying polymorphic functions) are available in examples for concision. The semantics we assume also directly follows \cite{pfpl}, so we also omit the details of this core here.  Only three new  kinds are needed for the SL to serve its role as the language used to   control typing and translation of the EL: $\kty$ (Sec. \ref{types}), $\kity$ (Sec. \ref{sec:type-translations}) and $\kitm$ (Sec. \ref{sec:introop}).

@@

\inferrule[k-otherty]{
    \sofkX{\sttyidx}{\keyw{Nat} \times \kity}
}{
    \sofkX{\sty{\keyw{other}[m]}{\sttyidx}}{\kty}
}\vspace{-10px}
\end{mathpar}
\caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
\label{fig:types}\vspace{-10px}
\end{figure*}

@@

\paragraph{Examples} Two examples of tycon contexts (each defining only one tycon)  and their signatures are shown in Figure \ref{fig:example-tycons}. \footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.}

Our first example, $\tcvar{rstr}$, has index kind $\krx$, which we assume classifies static regular expression patterns (defined as an inductive sum kind in the usual way). Types constructed by $\tcvar{rstr}$ will classify terms that behave as \emph{regular strings}, statically obeying the invariant that they are in the regular language specified by the type index \cite{sanitation-psp14}. For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.  The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably in libraries \cite{TSLs}. %We define the tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.

Our second example is the tycon $\tcvar{lprod}$, which will define a variant of labeled product type (labeled products are like record types, but maintain a row  ordering; record types are also definable in a manner discussed in the supplement, but maintaining an ordering simplifies our discussion). We choose the index kind of $\tcvar{lprod}$ to be $\klist{\klbl \times \kty}$, where list kinds are defined as inductive sums in the usual way, and $\klbl$ classifies static representations of row labels.  In the tycon context containing both tycon definitions, $\Phi_\text{rstr}\Phi_\text{lprod}$, we can define a labeled product type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$. Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and we again use kind-specific syntax, in this case for $\klist{\klbl \times \kty}$.

\subsection{Static Type Case Analysis}

\begin{figure}\small
\hspace{-10px}\begin{tabular}{ll}
$\begin{array}{l}\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\psi_\text{rstr}}{\\
\quad\tcstruct{\stx{rstr/trans}}{\\
\quad\tcstructn{\stx{rstr/intro}}};\\
\quad\keyw{syn}~\opname{conc} = \stx{rstr/conc};\\
\quad\keyw{syn}~\opname{case} = \stx{rstr/case};\\
\quad\cdots}\end{array}$ & $\begin{array}{l}\Phi_\text{lprod} := \tcdef{\tcvar{lprod}}{\psi_\text{lprod}}{\\
\quad\tcstruct{\stx{lprod/trans}}{\\
\quad\tcstructn{\stx{lprod/intro}}};\\
\quad\keyw{syn}~\opname{\#} = \stx{lprod/prj};\\
\quad\keyw{syn}~\opname{conc} = \stx{lprod/conc};\\
\quad\cdots}\end{array}$
\end{tabular}\vspace{3px}\\
$\psi_\text{rstr} := \tcsig{\krx}{\introsig{\keyw{Str}}; \opname{conc}[\kunit]; \opname{case}[\keyw{StrPattern}]; ...}$\\
$\psi_\text{lprod} := \tcsig{\klist{\klbl\times\kty}}{\introsig{\klist{\klbl}}; \opname{\#}[\klbl]; \opname{conc}[\kunit]; ...}$
\caption{Example tycon contexts and signatures.}
\label{fig:example-tycons}\vspace{-8px}
\end{figure}

@@

The dynamics (see supplement) are straightforwardly consistent with the intuition above.% We will see an example of its use in Sec. \ref{sec:targops}.

\subsection{Tycon Context Well-Definedness}

% \begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}
% \caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
% \label{fig:tycon-ctxs}
% \vspace{-8px}
% \end{figure}
The tycon context well-definedness judgement, $\vdash \Phi$, requires that all tycon names are unique and performs three additional checks,  described below (we omit (tcc-emp)).
\begin{mathpar}
\small
\inferrule[tcc-ext]{
    \vdash \Phi\\
    \tc \notin \text{dom}(\Phi)\\
    \keq{\emptyset}{\ktyidx}\\
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
}{
    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
}
\end{mathpar}
\subsection{Type Equivalence}\label{sec:type-equivalence}
The first check simplifies the handling of type equivalence: type index kinds must be \emph{equality kinds}, i.e. those for which semantic equivalence implies syntactic equality at normal form. We define these by the judgement $\keq{\kDelta}{\kappa}$ (see supplement). 
Equality kinds are similar to equality types as found in Standard ML \cite{Tofte:89:TheDefinitionOfStandardML}. The main implication of this choice is that type indices cannot contain static functions.

@@

Applying the substitutions $\gamma_\text{ex/abs}$ and $\delta_\text{ex/abs}$ in the conclusion of the rule, we arrive at the actual term translation $\iota_\text{ex}$, defined previously. Note that $\iota_\text{ex}$ has type $\tau_\text{paper}$ under the translation of $\Upsilon_\text{ex}$, i.e. $\vdash \Upsilon_\text{ex} \leadsto \Gamma_\text{ex}$ where $\Gamma_\text{ex} := title : \keyw{str}$.
